### 2025-05-14 docs/Homeownership_BankOfMomAndDad.tex
Improve table titile
Ready for submission

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Study of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 14 percentage points (29\%) of the homeownership rate among young households. In one adaption of my model, I quantify the contribution of parental transfers to the Black-White homeownership gap of 50\% among young adults, following the  methodology of \cite{Ashman2020} and \cite{aliprantis2022dynamics}. I find that parental transfers account for 7 percentage points (14\%) of the Black-White homeownership gap.

@@

I then study how illiquidity affects strategic behavior in the absence of commitment. As an illustration, consider a child who buys a house and brings no liquid wealth to the next period. If the parent does not transfer in the next period, the child will either face very low consumption—implying a high marginal value of wealth—or be forced to liquidate the house and incur sales costs. The parent dislikes both outcomes and may find it optimal to transfer enough to keep the child in the home. More precisely, children with wealthy parents use the sales costs as a commitment device to maintain low liquid wealth in the next period, raising their next-period marginal utility of wealth and thereby strengthening the parent’s transfer motive. As a result, 27\% of young households prefer illiquid housing over liquid housing. That parental wealth generates a preference for illiquidity is a new theoretical mechanism---distinct from existing explanations such as temptation or time inconsistency \citep[e.g.,][]{attanasio2024temptation,laibson1997golden}.

@@

I then examine the interplay  of transfers and housing for marginal propensities to consume (MPC). A substantial literature has sought to reconcile structural models with the high MPCs observed in empirical data.  In my model, the introduction of altruistic transfers lowers average MPCs by about 30\%, implying that standard life-cycle models may understate the discrepancy.  Moreover, one common explanation for high MPCs is that some households are wealthy but liquidity constrained---such as the ``wealthy hand-to-mouth'' in \cite{kaplan2022marginal}---but I find that ignoring parental transfers overstates their MPCs by about 50\%. Another explanation for high MPCs is heterogeneity in preferences---such as impatience \citep[e.g.,][]{aguiar2024hand}---yet such behavior is observationally similar to that of households with wealthy, altruistic parents.

@@

Two results stand out. First, parental wealth is at least as predictive as household net worth at all horizons, although the difference is never statistically significant. Second, the effect of current parental wealth grows with the time horizon. For instance, a 10\% increase in parental wealth raises the probability of retaining ownership after eight years by 0.45 percentage points among first-time owners. I return to these results later as part of the model validation.

@@

I first limit the sample to first-time buyers to focus on distress among new owners. In a univariate regression, a 10\% increase in parental wealth is associated with a 0.07 percentage point decrease in the probability of being behind on mortgages two years later ($p<0.1$) With the full set of controls, the effect increases slightly, though it is no longer statistically significant. These magnitudes are economically meaningful since 1.6\% of first-time owners are behind on their mortgage.In Columns 3 to 5, I expand the analysis to include all owners who are currently not behind, increasing the sample sizes and allowing us to include household fixed effects. In these specifications, a 10\% increase in parental wealth is associated with a 0.09 percentage point decrease in delinquency risk in the univariate OLS regression ($p<0.001$), a 0.02 percentage point decrease in the multivariate OLS regression (not significant), and a 0.07 percentage point decrease in the fixed effects regression ($p<0.1$).

@@

The exercise follows \cite{Chetty2007} closely. The outcome of interest is the change in log housing consumption, which is set to zero for households who do not move. For movers, housing consumption is defined as yearly rent while renting and as imputed rent for owners, set to 5\% of the market value \citep{Davis2008}. The sample is limited to household heads who experience only one unemployement spell between ages 25 and 44. I divide the sample by whether a household's parents were in the top parental wealth quartile at the time of unemployment and compare the two groups' housing consumption growth at unemployment. The results are displayed in Figure \ref{fig:housinggrowthrates}. Among households with non-wealthy parents, housing consumption falls by about 5\% during unemployment ($p<0.05$). In contrast, households with wealthy parents see a slight increase (not significant)in housing consumption during unemployment.\footnote{In Appendix \ref{app:eventstudy_details}, I also report the results of a formal event study using the same set of controls as in the previous regressions.}  I return to these results later as part of the model validation.

@@

I estimate the altruism parameter $\eta$ because, relative to previous research, the inclusion of housing introduces new transfer motives, requiring re-estimation to match observed intra-family transfers. Similarly, I estimate the homeownership utility $\chi$, since transfers generate new ownership motives. Finally, I estimate the minimum size of owner-occupied housing, $h_0$, a key parameter in models of homeownership, because housing choices interact with transfer behavior.

@@

First, with a calibrated discount factor of $\beta=0.92$, the model matches the wealth levels of both young and old households. It also replicates the correlation between parental wealth and children's homeownership: in the PSID, young homeowners' parents are 2.5 times wealthier than young renters' parents, compared to about 2 times in the model. This parental wealth gradient effectively captures the intergenerational link in homeownership.

Second, the model aligns well with other moments related to homeownership. It slightly overshoots the age of first homeownership and overpredicts the aggregate homeownership rate of 65\%, primarily due to an excess of homeowner parents stemming from their lack of income risk (see Appendix \ref{sec:robust_incomerisk}). Mortgages are slightly smaller than in the data, as households repay and re-borrow flexibly. The model also matches the LTV ratio for buyers well.\footnote{The empirical LTV is measured at the first observation of ownership, not necessarily at origination, so it can reflect price appreciation, improvements, and mortgage paydown.}

Third, the model also matches transfer-related moments well. It predicts average parental transfers of 2.3\% of parental wealth, close to the 2.6\% estimated from the 2013 Transfer Module (excluding outliers below 0\% or above 100\%). It accurately captures the share of households receiving transfers within two years of house purchases, suggesting proper synchronization of housing transactions and transfers. Renters also receive transfers more often than owners in both the model and the data, though the model slightly overstates this gap.

@@

\subsection{The Effect of Parental Transfers on Homeownership}
Table \ref{tab:modelfitnoa} reports how the moments change without altruism and transfers. Homeownership rates decline by 14 percentage points from 49\%, a decrease of 29\% for households aged 25 to 44. This is a large decrease, since households endogenously choose to save more without transfers. By comparison, homeownership only decreased by about 8 percentage points during the financial crisis.

The reduction in homeownership cannot be attributed to an inability to afford buying: Without altruism, median wealth of the young increases (increasing homeownership, in isolation) to the average wealth level at purchase with altruism. The main driver is that the ownership threshold shifts out in wealth without altruism, as we saw in the policy functions (Section \ref{sec:policyfuncs}). Without transfers then, the average wealth at purchase increases by about \$12,500 (about 33\%), the age of first ownership is delayed by about 2.5 years, and the LTV at purchase falls by 13 percentage points. Although homeownership of young households declines sharply, the overall homeownership rate of all households (ages 25 to 74) only falls by 6 percentage points (8\%). Hence, the main aggregate effect of parental transfers is that they induce earlier homeownership. Finally, without altruism, the parental wealth gradient in homeownership decreases from 1.94 to 1.14, with the remaining effect driven by the integenerational correlation in initial wealth and persistent labor productivity.

@@

\begin{table}
	\center
\begin{threeparttable}[tb]
				\singlespacing
		%			\footnotesize
		\caption{Black-White Homeownership Rate}\label{tab:BW_table}
		\input{../tabfig/postest/BW_mostrecent.tex}
		\footnotesize
		\textit{Notes:} Wealth is measured in 1000s of 2016 US dollars. See main text for details.
	\end{threeparttable}
\end{table}

@@

The observed Black-White homeownership gap is $\frac{0.54 - 0.27}{0.54} = 50\%$, compared to 52\% in the model with altruism. Removing altruism reduces the gap to 45\%, implying that parental transfers account for 7 percentage points---or 14\%---of the gap. This is a sizeable effect, especially given that the Black–White homeownership gap has remained largely unchanged since the passage of the Fair Housing Act in 1968. While Black and White households receive transfers at similar rates in the model, transfers are less influential for Black households, who are further from the ownership threshold and receive smaller amounts. These results highlight the importance of racial differences in parental income, which shape both wealth accumulation and transfer capacity.

@@

Table \ref{tab:quant_endogenprices} reports the results. Allowing house prices to adjust does not meaningfully decrease the contribution of transfers to homeownership or other aggregate outcomes. This is due to the small difference in homeownership rates (ages 25 to 74) with and without altruism. Prices decline by just 7\% even with a supply elasticity as low as 1. Thus, transfers account for between 10 and 13 percentage points of the homeownership rate, depending on the supply elasticity.

@@

\textit{Relaxing Mortgaging Regulation:} To study how parental transfers interact with down payment requirements, I tighten the LTV limit by 3 percentage points. The tighter limit decreases homeownership among the young by 5 percentage points (10\%). Homeownership declines more among households with parents in the middle of the wealth distribution, reducing the parental wealth gradient. While initially counterintuitive, these middle-wealth households rely heavily on small transfers to overcome borrowing constraints; poorer households remain constrained, and richer ones are unaffected. Thus, looser borrowing constraints not only boost homeownership overall but also amplify the importance of parental wealth. Consistent with this mechanism, \citet{wold2024housing} find that owners with poorer parents have lower LTV ratios at origination.

\textit{Lowering PMI:} To study how parental transfers interact with the PMI, which increases the interest rate on high LTV loans, I relax the PMI limit by 7 percentage points. The relaxed PMI limit increases homeownership by 2 percentage points (4\%). Similar to the LTV regulation, homeownership among households with parents in the middle of the wealth distribution rises more, further widening the parental wealth gradient. This result suggests that relaxing the PMI limit increases not only homeownership for all households, but also the role of parental wealth.

\textit{Lowering House Prices:} To study how parental transfers interact with housing affordability, I lower house prices by \$6,000 (7\%). As prices decline, homeownership increases for all households, though more so for households with the richest parents, slightly increasing the role of parental wealth.

\textit{Reducing Purchase Costs:} To study how closing costs and first-time buyer assistance interact with parental transfers, I lower the purchase cost ($m_b$) by 2 percentage points---equivalent to a grant of about \$6,250.\footnote{Many U.S. states offer closing cost assistance to first-time buyers. In England, stamp duty was reduced for first-time buyers in 2017.} While homeownership rises, the gains are concentrated among households with wealthy parents, amplifying the parental wealth gradient. Lowering purchase costs has little effect for renters with non-wealthy parents, who remain constrained by LTV limits and sales costs. In contrast, for household with non-wealthy parents, lower purchase costs reduce the effective price—the main remaining barrier to ownership. This suggests that grants or credits for first-time buyers primarily benefit households of the wealthiest.

\textit{Reducing Sales Costs:} 
The earlier empirical results suggested that parental wealth helps offset the illiquidity of housing. To study this interaction, I lower the sales cost $m_s$ from \parms\% to 5. Strikingly, the homeownership rate decreases by 2 percentage points (4\%). This counterintuitive result reflects selection into ownership. homeownership rises slightly among households with the wealthiest and poorest parents but drops sharply---by 9 percentage points---among households with moderately wealthy parents. For these households, lower sales costs weaken the ability to use housing as a commitment device to secure future transfers. In contrast, children of the poorest parents (who rarely receive transfers) and the wealthiest parents (who expect transfers regardless) are less affected.

@@

Without altruism, all households prefer to remove sales costs. With altruism (baseline calibration), I find that while all parents prefer to remove sales costs, roughly one-quarter of children prefer them. The key determinant is parental wealth: Parents of children who prefer sales costs are approximately three times wealthier. Children who prefer sales costs are also more likely to receive transfers (48\% vs. 13\%) and, conditional on receiving transfers, receive three times larger transfers. Notably, households that own their homes are more likely to prefer sales costs on their housing.

@@

To explore why some households prefer illiquidity, Figure~\ref{fig:prefliq} breaks the sample by liquidity preference and age. Figure~\ref{fig:pref} shows that illiquidity preference declines with age, consistent with reduced reliance on parental transfers; around 50\% of the youngest households prefer illiquidity. The next panels plot homeownership, parental wealth, and child wealth by liquidity preference. Since illiquidity can only be used to extract transfers when the household owns a home and the parent is wealthy, those who prefer it have higher homeownership rates and wealthier parents. Figure~\ref{fig:pref_wealthk} shows that illiquidity-preferring households are initially wealthier, but this reverses around age 33. This flip reflects a tension: to use housing as a commitment device, households must be wealthy enough to own, yet poor enough to remain liquidity-constrained and elicit transfers.

@@

\subsection{MPCs}
If altruism makes illiquid assets more attractive, a natural question is how altruism affects MPCs, as liquidity constraints are an important driver of MPCs \citep[see e.g.,][]{aguiar2024hand,Kaplan2014,fagereng2021mpc}. We begin by computing counterfactual MPCs for households in the stationary distribution, assuming no altruistic transfers. The average MPC of young households is 0.35. The MPC of liquidity constrained households (0.44) is about three times that of unconstrained households (0.16). While these MPCs are lower than empirical estimates---typically ranging from 0.5 to 0.9---they are relatively standard in models not designed to generate high MPCs.

@@

\begin{table}[tb]
	\center\singlespacing	
	\begin{threeparttable}[tb]
		\caption{Altruistic Transfers Decrease the MPC}
		\input{../tabfig/postest/MPCs.tex}
	\end{threeparttable}
	{\begin{footnotesize}\begin{flushleft}\vspace{-0.1in}
		\textit{Notes:} Sample includes all households aged 25-44 in the stationary distribution. Following \cite{Kaplan2014}, households are liquidity constrained if they have negative liquid wealth $(b_k)$ exceeding 16.5\% of annual earnings. The MPC is calculated from a \$5,000 positive shock.\end{flushleft}\end{footnotesize}}			
\end{table}

@@

Using a counterfactual experiment without altruism (a standard life-cycle model), I find that transfers account for 14 percentage points (29\%) of the homeownership rate. I show that policies that lower entry barriers to homeownership amplify the importance of parental wealth on housing outcomes, while those that lower the downsides of illiquid homeownership reduce the importance of parental wealth. Finally, I study the role of housing illiquidity and parental transfers. My paper thus contributes to a better understanding of the determinants of homeownership and to the growing literature on the role of family economics in shaping macroeconomic outcomes (e.g., \cite{Doepke2016a,Daruich2018}).

@@

The results are reported in Table \ref{tab:robust}. We see that the introduction of aggregate price uncertainty increases savings for old households, with and without altruism. It also induces households to buy housing slightly later, with lower LTV's and higher wealth at purchase. However, the main quantitative finding remains: Parental transfers now account for 11pp (25\%) of the homeownership rate. I have also perform robustness checks where I allow for transitory idiosyncratic dividend/return shocks as in \cite{Chang2024} for homeowners, and find that, as expected, idiosyncratic price risk have even smaller quantitative effects.

@@

\section{Numerical Details}\label{sec:computational}
I now briefly discuss some details in the numerical solution of the model. Due to the nonconvex nature of the decision problems---due to discrete housing choices, noncontinuous transfer choices, and occasionally binding constraints---I use grid search. That is, I define a grid of possible choices for each choice variable, and for each state I loop over all possible choices, and the optimal policy is the one that maximizes the utility. The value and policy functions are linearly interpolated over child and parental wealth, conditional on the discrete choices. This solution algorithm satisfies both criteria established by \cite[p.30]{Barczyk2020}: ``algorithms should be able to deal with locally convex and even discontinuous value functions''. Naturally, the solution accuracy improves with denser grids. I use {\parnstate} nodes in the state vectors and {\parnchoice} nodes in choice variable grids.

@@

Intuitively, this cyclical behavior can arise from the interaction between the discrete rent-or-own choice and strategic interactions. For example, a renting child may decide to buy given the parent's policy functions at a certain state. In the next iteration, the parent observes this and adjusts savings to discourage the child from buying, prompting the child to reoptimize. The cycle then repeats, with the child buying in one iteration and not buying in the next. When I solve the model without homeownership (reducing it to a standard altruistic consumption-savings model as in \cite{Barczyk2020a} and \cite{Chu2020}), the value function iterations consistently converge to a unique fixed point. Similarly, removing transaction costs, making housing more liquid and no longer a state variable, reduces the likelihood of cyclical fixed points. Denser grids also decreases the likelihood of cyclical behavior. In practice, the largest differences in policy functions within cycles are at the upper edges of the parental wealth grid. Since the upper edge is set so high that no households are close to it, this has little impact on the solution.

>>>>

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Study of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 13 percentage points (27\%) of the homeownership rate among young households. In one adaption of my model, I quantify the contribution of parental transfers to the Black-White homeownership gap of 50\% among young adults, following the  methodology of \cite{Ashman2020} and \cite{aliprantis2022dynamics}. I find that parental transfers account for 11 percentage points (20\%) of the Black-White homeownership gap.

@@

I then study how illiquidity affects strategic behavior in the absence of commitment. As an illustration, consider a child who buys a house and brings no liquid wealth to the next period. If the parent does not transfer in the next period, the child will either face very low consumption—implying a high marginal value of wealth—or be forced to liquidate the house and incur sales costs. The parent dislikes both outcomes and may find it optimal to transfer enough to keep the child in the home. More precisely, children with wealthy parents use the sales costs as a commitment device to maintain low liquid wealth in the next period, raising their next-period marginal utility of wealth and thereby strengthening the parent’s transfer motive. As a result, 22\% of young households prefer illiquid housing over liquid housing. That parental wealth generates a preference for illiquidity is a new theoretical mechanism---distinct from existing explanations such as temptation or time inconsistency \citep[e.g.,][]{attanasio2024temptation,laibson1997golden}.

@@

I then examine the interplay  of transfers and housing for marginal propensities to consume (MPC). A substantial literature has sought to reconcile structural models with the high MPCs observed in empirical data.  In my model, the introduction of altruistic transfers lowers average MPCs by about 20\%, implying that standard life-cycle models may understate the discrepancy.  Moreover, one common explanation for high MPCs is that some households are wealthy but liquidity constrained---such as the ``wealthy hand-to-mouth'' in \cite{kaplan2022marginal}---but I find that ignoring parental transfers overstates their MPCs by about 30\%. Another explanation for high MPCs is heterogeneity in preferences---such as impatience \citep[e.g.,][]{aguiar2024hand}---yet such behavior is observationally similar to that of households with wealthy, altruistic parents.

@@

Two results stand out. First, parental wealth is at least as predictive as household net worth at all horizons, although the difference is never statistically significant. Second, the effect of current parental wealth grows with the time horizon. For instance, a 10\% increase in parental wealth raises the probability of retaining ownership after eight years by about 0.45 percentage points among first-time owners. I return to these results later as part of the model validation.

@@

I first limit the sample to first-time buyers to focus on distress among new owners. In a univariate regression, a 10\% increase in parental wealth is associated with a 0.07 percentage point decrease in the probability of being behind on mortgages two years later ($p<0.1$) With the full set of controls, the effect increases slightly, though it is no longer statistically significant. These magnitudes are economically meaningful since 1.6\% of first-time owners are behind on their mortgage. In Columns 3 to 5, I expand the analysis to include all owners who are currently not behind, increasing the sample sizes and allowing us to include household fixed effects. In these specifications, a 10\% increase in parental wealth is associated with a 0.09 percentage point decrease in delinquency risk in the univariate OLS regression ($p<0.001$), a 0.02 percentage point decrease in the multivariate OLS regression (not significant), and a 0.07 percentage point decrease in the fixed effects regression ($p<0.1$).

@@

The exercise follows \cite{Chetty2007} closely. The outcome of interest is the change in log housing consumption, which is set to zero for households who do not move. For movers, housing consumption is defined as yearly rent while renting and as imputed rent for owners, set to 5\% of the market value \citep{Davis2008}. The sample is limited to household heads who experience only one unemployement spell between ages 25 and 44. I divide the sample by whether a household's parents were in the top parental wealth quartile at the time of unemployment and compare the two groups' housing consumption growth at unemployment. The results are displayed in Figure \ref{fig:housinggrowthrates}. Among households with non-wealthy parents, housing consumption falls by about 5\% during unemployment ($p<0.05$). In contrast, households with wealthy parents see a slight increase (not significant) in housing consumption during unemployment. As shown in Appendix~\ref{app:eventstudy_details}, the main finding---that households with wealthy parents do not downsize during unemployment---is also retained in a formal event study that includes the same set of controls as the previous regressions. I return to these results later as part of the model validation.

@@

I estimate the altruism parameter $\eta$ because, relative to previous research, the inclusion of housing introduces new transfer motives, requiring re-estimation to match observed intra-family transfers. Similarly, I estimate the homeownership utility $\chi$, since transfers generate new ownership motives. Finally, I estimate the relative size of owner-occupied housing, $h_0$, a key parameter in models of homeownership, because housing choices interact with transfer behavior.

@@

First, with a calibrated discount factor of $\beta=0.92$, the model matches the wealth levels of both young and old households. It also replicates the correlation between parental wealth and children's homeownership: in the PSID, young homeowners' parents are 2.6 times wealthier than young renters' parents, compared to about 2 times in the model. This parental wealth gradient effectively captures the intergenerational link in homeownership.

Second, the model aligns well with other moments related to homeownership. It slightly overshoots the age of first homeownership and overpredicts the aggregate homeownership rate of 64\%, primarily due to an excess of homeowner parents stemming from their lack of income risk (see Appendix \ref{sec:robust_incomerisk}). Mortgages are slightly smaller than in the data, as households repay and re-borrow flexibly. The model also matches the LTV ratio for buyers well.\footnote{The empirical LTV is measured at the first observation of ownership, not necessarily at origination, so it can reflect price appreciation, improvements, and mortgage paydown.}

Third, the model also matches transfer-related moments well. It correctly predicts average parental transfers of 2.6\% of parental wealth, as estimated from the 2013 Transfer Module (excluding outliers below 0\% or above 100\%). It roughly matches the share of households receiving transfers within two years of house purchases, suggesting proper synchronization of housing transactions and transfers. Renters also receive transfers more often than owners in both the model and the data, though the model slightly overstates this gap.

@@

\subsection{The Effect of Parental Transfers on Homeownership}
Table \ref{tab:modelfitnoa} reports how the moments change without altruism and transfers. Homeownership rates decline by 13 percentage points from 48\%, a decrease of 27\% for households aged 25 to 44. This is a large decrease, since households endogenously choose to save more without transfers. By comparison, homeownership only decreased by about 8 percentage points during the financial crisis.

The reduction in homeownership cannot be attributed to an inability to afford buying: Without altruism, median wealth of the young increases (increasing homeownership, in isolation). The main driver is that the ownership threshold shifts out in wealth without altruism, as we saw in the policy functions (Section \ref{sec:policyfuncs}). Without transfers then, the average wealth at purchase increases by about 25\%, the age of first ownership is delayed by about 2.5 years, and the LTV at purchase falls by 12 percentage points. Although homeownership of young households declines sharply, the overall homeownership rate of all households (ages 25 to 74) only falls by 6 percentage points (9\%). Hence, the main aggregate effect of parental transfers is that they induce earlier homeownership. Finally, without altruism, the parental wealth gradient in homeownership decreases from 2.0 to 1.15, with the remaining effect driven by the integenerational correlation in wealth and income. Hence, the positive correlation between parental wealth and housing outcomes is almost entirely driven by parental transfers.

@@

\begin{table}
	\center
\begin{threeparttable}[tb]
				\singlespacing
		%			\footnotesize
		\caption{Altruism and the Black-White Homeownership Gap}\label{tab:BW_table}
		\input{../tabfig/postest/BW_mostrecent.tex}
		\footnotesize
		\textit{Notes:} Wealth is measured in 1000s of 2016 US dollars. See main text for details.
	\end{threeparttable}
\end{table}

@@

The observed Black-White homeownership gap is $\frac{0.53 - 0.26}{0.53} = 51\%$, compared to 54\% in the model with altruism. Removing altruism reduces the gap to 43\%, implying that parental transfers account for 11 percentage points---or 20\%---of the gap. This is a sizeable effect, especially given that the Black–White homeownership gap has remained largely unchanged since the passage of the Fair Housing Act in 1968. While Black and White households receive transfers at similar rates in the model, transfers are less influential for Black households, who are further from the ownership threshold and receive smaller amounts. These results highlight the importance of racial differences in parental income, which shape both wealth accumulation and transfer capacity.

@@

Table \ref{tab:quant_endogenprices} reports the results. Allowing house prices to adjust does not meaningfully decrease the contribution of transfers to homeownership or other aggregate outcomes. This is due to the small difference in homeownership rates (ages 25 to 74) with and without altruism. Prices decline by just 5\% even with a supply elasticity as low as 1. Thus, transfers account for between 10 and 13 percentage points of the homeownership rate, depending on the supply elasticity.

@@

\textit{Relaxing Mortgaging Regulation:} To study how parental transfers interact with down payment requirements, I tighten the LTV limit by 3 percentage points. The tighter limit decreases homeownership among the young by 7 percentage points (15\%). Homeownership declines more among households with parents in the middle of the wealth distribution, reducing the parental wealth gradient. While initially counterintuitive, these middle-wealth households rely heavily on small transfers to overcome borrowing constraints; poorer households remain constrained, and richer ones are unaffected. Thus, looser borrowing constraints not only boost homeownership overall but also amplify the importance of parental wealth. Consistent with this mechanism, \citet{wold2024housing} find that owners with poorer parents have lower LTV ratios at origination.

\textit{Lowering PMI:} To study how parental transfers interact with the PMI, which increases the interest rate on high LTV loans, I relax the PMI limit by 7 percentage points. The relaxed PMI limit increases homeownership by 1 percentage points (2\%). Similar to the LTV regulation, homeownership among households with parents in the middle of the wealth distribution rises more, further widening the parental wealth gradient. This result suggests that relaxing the PMI limit increases not only homeownership for all households, but also the role of parental wealth.

\textit{Lowering House Prices:} To study how parental transfers interact with housing affordability, I lower house prices by 2\%. As prices decline, homeownership increases for all households, though more so for households with the richest parents, increasing the role of parental wealth.

\textit{Reducing Purchase Costs:} To study how closing costs and first-time buyer assistance interact with parental transfers, I lower the purchase cost ($m_b$) by 2 percentage points---equivalent to a grant of about \$6,000.\footnote{Many U.S. states offer closing cost assistance to first-time buyers. In England, stamp duty was reduced for first-time buyers in 2017.} While homeownership rises, the gains are concentrated among households with wealthy parents, amplifying the parental wealth gradient. Lowering purchase costs has little effect for renters with non-wealthy parents, who remain constrained by LTV limits and sales costs. In contrast, for household with non-wealthy parents, lower purchase costs reduce the effective price—the main remaining barrier to ownership. This suggests that grants or credits for first-time buyers primarily benefit households with richer parents.

\textit{Reducing Sales Costs:} 
The earlier empirical results suggested that parental wealth helps offset the illiquidity of housing. To study this interaction, I lower the sales cost $m_s$ by 2 percentage points. Strikingly, the homeownership rate decreases by 3 percentage points (6\%). This counterintuitive result reflects selection into ownership. Homeownership rises slightly for most households but drops sharply---by 6 percentage points---among households with moderately wealthy parents. For these households, lower sales costs weaken the ability to use housing as a commitment device to secure future transfers. In contrast, children of the poorest parents (who rarely receive transfers) and the wealthiest parents (who expect transfers regardless) are less affected.

@@

Without altruism, all households prefer to remove sales costs. With altruism (baseline calibration), I find that while all parents prefer to remove sales costs, roughly one-quarter of children prefer them. The key determinant is parental wealth: Parents of children who prefer sales costs are approximately three times wealthier. Children who prefer sales costs are also more likely to receive transfers (51\% vs. 15\%) and, conditional on receiving transfers, receive three times larger transfers. Notably, households that own their homes are more likely to prefer sales costs on their housing.

@@

To explore why some households prefer illiquidity, Figure~\ref{fig:prefliq} breaks the sample by liquidity preference and age. Figure~\ref{fig:pref} shows that illiquidity preference declines with age, consistent with reduced reliance on parental transfers; around 40\% of the youngest households prefer illiquidity. The next panels plot homeownership, parental wealth, and child wealth by liquidity preference. Since illiquidity can only be used to extract transfers when the household owns a home and the parent is wealthy, those who prefer it have higher homeownership rates and wealthier parents. Figure~\ref{fig:pref_wealthk} shows that illiquidity-preferring households are initially wealthier, but this reverses around age 33. This flip reflects a tension: to use housing as a commitment device, households must be wealthy enough to own, yet poor enough to remain liquidity-constrained and elicit transfers.

@@

\subsection{MPCs}
If altruism makes illiquid assets more attractive, a natural question is how altruism affects MPCs, as liquidity constraints are an important driver of MPCs \citep[see e.g.,][]{aguiar2024hand,Kaplan2014,fagereng2021mpc}. We begin by computing counterfactual MPCs for households in the stationary distribution, assuming no altruistic transfers. The average MPC of young households is 0.32. The MPC of liquidity constrained households (0.43) is about three times that of unconstrained households (0.14). While these MPCs are lower than empirical estimates---typically ranging from 0.5 to 0.9---they are relatively standard in models not designed to generate high MPCs.

@@

\begin{table}[tb]
	\center\singlespacing	
	\begin{threeparttable}[tb]
		\caption{Altruistic Transfers Decrease the MPC}
		\input{../tabfig/postest/MPCs.tex}
	\end{threeparttable}
	{\begin{footnotesize}\begin{flushleft}\vspace{-0.1in}
		\textit{Notes:} The sample includes all households aged 25-44 in the stationary distribution. Following \cite{Kaplan2014}, households are liquidity constrained if they have negative liquid wealth $(b_k)$ exceeding 16.5\% of annual earnings. The MPC is calculated from a \$5,000 positive shock. To calculate the MPCs with future transfers only, we use the policy functions with altruism (baseline) but simulate households after unexpectedly confiscating any transfers. To calculate the MPCs with current transfers only, we use the parental policy function with altruism (baseline) and the children’s policy function without altruism. The children then receive the transfers but behave as if there are no future transfers. \end{flushleft}\end{footnotesize}}			
\end{table}

@@

Using a counterfactual experiment without altruism (a standard life-cycle model), I find that transfers account for 13 percentage points (27\%) of the homeownership rate. I show that policies that lower entry barriers to homeownership amplify the importance of parental wealth on housing outcomes, while those that lower the downsides of illiquid homeownership reduce the importance of parental wealth. Finally, I study the role of housing illiquidity and parental transfers. My paper thus contributes to a better understanding of the determinants of homeownership and to the growing literature on the role of family economics in shaping macroeconomic outcomes (e.g., \cite{Doepke2016a,Daruich2018}).

@@

The results are reported in Table \ref{tab:robust}. We see that the introduction of aggregate price uncertainty increases savings for old households, with and without altruism. It also induces households to buy housing slightly later, with lower LTV's and higher wealth at purchase. However, the main quantitative finding remains: Parental transfers now account for 10pp (23\%) of the homeownership rate. I have also perform robustness checks where I allow for transitory idiosyncratic dividend/return shocks as in \cite{Chang2024} for homeowners, and find that, as expected, idiosyncratic transitory price risk have even smaller effects.

@@

\section{Numerical Details}\label{sec:computational}
I now briefly discuss some details in the numerical solution of the model. Due to the nonconvex nature of the decision problems---due to discrete housing choices, noncontinuous transfer choices, and occasionally binding constraints---I use grid search. That is, I define a grid of possible choices for each choice variable, and for each state I loop over all possible choices, and the optimal policy is the one that maximizes the utility. The value and policy functions are linearly interpolated over child and parental wealth, conditional on the discrete choices. This solution algorithm satisfies both criteria established by \cite[p.30]{Barczyk2020}: ``algorithms should be able to deal with locally convex and even discontinuous value functions.'' Naturally, the solution accuracy improves with denser grids. I use {\parnstate} nodes in the state vectors and {\parnchoice} nodes in choice variable grids.

@@

Intuitively, this cyclical behavior can arise from the interaction between the discrete rent-or-own choice and strategic interactions. For example, a renting child may decide to buy given the parent's policy functions at a certain state. In the next iteration, the parent observes this and adjusts savings to discourage the child from buying, prompting the child to reoptimize. The cycle then repeats, with the child buying in one iteration and not buying in the next. When I solve the model without homeownership (reducing it to a standard altruistic consumption-savings model as in \cite{Barczyk2020a} and \cite{Chu2020}), the value functions always converge to a unique fixed point. Similarly, removing transaction costs, making housing more liquid and no longer a state variable, reduces the likelihood of cyclical fixed points. Denser grids also decreases the likelihood of cyclical behavior. In practice, the largest differences in policy functions within cycles are at the upper edges of the parental wealth grid. Since the upper edge is set so high that no households are close to it, this has little impact on the solution.
